\section{DAC Unit}

The digital/analog unit works with the two-channel \gls{DAC} hardware peripheral. It can be used in two modes: \gls{DC} output, and waveform generation.

The waveform mode implements direct digital synthesis (explained in section~\ref{sec:theory-dac-dds}) of a sine, rectangle, sawtooth or triangle wave. The generated frequency can be set with a sub-hertz precision up to the lower tens of kHz. The two outputs can use a different waveform shape, can be synchronized, and their phase offset and frequency are dynamically adjustable.

\subsection{DAC Configuration}

\begin{inicode}
[DAC:dac@13]
# Enabled channels (1:A4, 2:A5)
ch1_enable=Y
ch2_enable=Y
# Enable output buffer
ch1_buff=Y
ch2_buff=Y
# Superimposed noise type (NONE,WHITE,TRIANGLE) and nbr. of bits (1-12)
ch1_noise=NONE
ch1_noise-level=3
ch2_noise=NONE
ch2_noise-level=3
\end{inicode}

\subsection{DAC Commands}

Channels are specified in all commands as a bit map:

\begin{itemize}[nosep]
	\item 0x01 - channel 1
	\item 0x02 - channel 2
	\item 0x03 - both channels affected at once
\end{itemize}

\begin{cmdlist}
	0 & \cname{WAVE\_DC}
	Set a \gls{DC} level, disable \gls{DDS} for the channel
	& \begin{cmdreq}
		\cfield{u8} channels
		\cfield{u16} level (0--4095)
	\end{cmdreq} \\

	1 & \cname{WAVE\_SINE}
	Start a sine waveform
	& \begin{cmdreq}
		\cfield{u8} channels
	\end{cmdreq} \\

	2 & \cname{WAVE\_TRIANGLE}
	Start a symmetrical triangle waveform
	& \begin{cmdreq}
		\cfield{u8} channels
	\end{cmdreq} \\

	3 & \cname{WAVE\_SAWTOOTH\_UP}
	Start a rising sawtooth waveform
	& \begin{cmdreq}
		\cfield{u8} channels
	\end{cmdreq} \\

	4 & \cname{WAVE\_SAWTOOTH\_DOWN}
	Start a dalling sawtooth waveform
	& \begin{cmdreq}
		\cfield{u8} channels
	\end{cmdreq} \\

	5 & \cname{WAVE\_RECTANGLE}
	Start a rectangle waveform
	& \begin{cmdreq}
		\cfield{u8} channels
		\cfield{u16} on-time (0--8191)
		\cfield{u16} high level (0--4095)
		\cfield{u16} low level (0--4095)
	\end{cmdreq} \\

	10 & \cname{SYNC}
	Synchronize the two channels. The phase accumulator is reset to zero.
	& \\

	20 & \cname{SET\_FREQUENCY}
	Set the channel frequency
	& \begin{cmdreq}
		\cfield{u8} channels
		\cfield{float32} frequency
	\end{cmdreq} \\

	21 & \cname{SET\_PHASE}
	Set a channel's phase. It is recommended to only set the phase of one channel, leaving the other at 0°.
	& \begin{cmdreq}
		\cfield{u8} channels
		\cfield{u16} phase (0--8191)
	\end{cmdreq} \\

	22 & \cname{SET\_DITHER}
	Control the dithering function of the \gls{DAC} block. A high noise amplitude can cause an overflow to the other end of the output range due to a bug in the \gls{DAC} peripheral. Use value 255 to leave the parameter unchanged.

	& \begin{cmdreq}
		\cfield{u8} channels
		\cfield{u8} noise type (0-none, 1-white, 2-triangle)
		\cfield{u8} number of noise bits (1--12)
	\end{cmdreq} \\
\end{cmdlist}



