% Abstrakt

\begin{abstract-english}
This thesis documents the development of a general purpose software and hardware platform for interfacing low level hardware from high level programming languages and applications run on a PC, using USB and also wirelessly.

The requirements of common engineering tasks and problems occurring in the university environment were evaluated to design an extensible, reconfigurable hardware module that would make a practical, versatile, and low cost tool that in some cases also eliminates the need for professional measurement and testing equipment.

Several hardware prototypes and control libraries in programming languages C and Python have been developed. The Python library additionally integrates with MATLAB scripts. The devices provide access to a range of hardware buses and low level features and can be reconfigured using configuration files stored inside its permanent memory.
\end{abstract-english}

\begin{abstract-czech}
\begin{otherlanguage}{czech}
Tato práce popisuje vývoj univerzální softwarové a~hardwarové platformy pro přístup k~hardwarovým sběrnicím a~elektrickým obvodům z~prostředí vysokoúrovňových programovacích jazyků a~aplikací běžících na PC, a~to za využití USB a~také bezdrátově.

Byly vyhodnoceny požadavky typických problémů, vyskytujících se v~praxi při práci s~vestavěnými systémy a~ve výuce, pro návrh snadno rozšiřitelného a~přenastavitleného hardwarového modulu který bude praktickým, pohodlným a~dostupným nástrojem který navíc v~některých případech může nahradit profesionální laboratorní přístroje.

Bylo navrženo několik prototypů hardwarových modulů, spolu s~obslužnými knihovnami v~jazycích C a~Python; k~modulu lze také přistupovat z~prostředí MATLAB. Přístroj umožňuje přístup k~většině běžných hardwarových sběrnic a~umožňuje také např. měřit frekvenci a~vzorkovat či generovat analogové signály.
\end{otherlanguage}
\end{abstract-czech}
