% Abstrakt

\begin{abstract-english}
This thesis documents the development of a universal software and hardware platform providing high-level user applications running on a PC with access to GPIO pins, hardware buses, and signal acquisition and generation functions, using a USB, USART, or wireless connection.

The requirements of common engineering tasks and problems occurring in the university environment were evaluated to design an extensible, reconfigurable hardware module that would make a versatile and low-cost tool that in some cases eliminates the need for professional measurement and testing equipment.

We designed custom hardware modules, an extensible microcontroller firmware, and PC support libraries for programming languages C and Python. The Python library may further  be used in MATLAB scripts. The devices provide access to hardware buses (\IIC, SPI, USART, and 1-Wire) and microcontroller peripherals (GPIO, ADC, and DAC), implement frequency measurement and other useful features. They are configured in INI files accessed through on a virtual USB mass storage device, or programmatically.
\end{abstract-english}

\begin{abstract-czech}
\begin{otherlanguage}{czech}
Tato práce popisuje vývoj univerzální softwarové a~hardwarové platformy pro přístup k~hardwarovým sběrnicím a~elektrickým obvodům z~prostředí vysokoúrovňových programovacích jazyků a~aplikací běžících na PC, a~to za využití USB, UARTU a~také bezdrátově.

Byly vyhodnoceny požadavky typických problémů vyskytujících se při práci s~vestavěnými systémy a~ve výuce za účelem návrhu snadno rozšiřitelného a~přenastavitleného hardwarového modulu který bude praktickým, pohodlným a~dostupným nástrojem, který navíc v~některých případech může nahradit profesionální laboratorní přístroje.

Bylo navrženo několik prototypů hardwarových modulů spolu s~obslužnými knihovnami v~jazycích C a~Python; Python knihovnu lze dále používat z~prostředí MATLAB. Navržené moduly umožňují přístup k~většině běžných hardwarových sběrnic (\IIC, SPI, USART, 1-Wire) a~umožňuje také např. měřit frekvenci a~vzorkovat či generovat analogové signály.
\end{otherlanguage}
\end{abstract-czech}
